The curious world of constructive mathematics

### 2022-09-02 algar-lecture-notes.tex
typo
Intepretation of statements

\begin{prop}There are irrational numbers~$x$ and~$y$ such that~$x^y$ is
rational.
\end{prop}
\begin{proof}[First proof] The number~$\sqrt{2}^{\sqrt{2}}$ is rational or
irrational. In the first case, set~$x \defeq \sqrt{2}$, $y \defeq \sqrt{2}$.
In the second case, set~$x \defeq \sqrt{2}^{\sqrt{2}}$, $y \defeq \sqrt{2}$.
\end{proof}
\begin{proof}[Second proof] Set~$x \defeq \sqrt{2}$ and~$y \defeq \log_{\sqrt{2}} 3$.
Then~$x^y = 3$ is rational. The verification that~$y$ is irrational is even
easier than that of~$\sqrt{2}$.\footnote{Let~$y = p/q$ with~$p, q \in \ZZ$
and~$q \neq 0$. Since~$y > 0$, we may assume~$p, q \in \NN$. Then~$3 =
(\sqrt{2})^{p/q}$, hence~$3^{2q} = 2^p$. This is in contradiction to the
uniqueness of the prime factor decomposition, since the factor~$3$ occurs on
the left but not on the right.}
\end{proof}

@@

\marginpar{\dbend}
In constructive mathematics, we do \emph{not} claim that the principle of
excluded middle is false. Indeed, intuitionistic logic is downwardly compatible
with classical logic (every intuitionistic proof is a fortiori also a classical
proof), and some special instances of the principle of excluded middle are
intuitionistically verifiable (an example is given in
Proposition~\ref{prop:discreteness}). Instead, in constructive mathematics we
merely do not use the principle of excluded middle.

>>>>

\begin{prop}There are irrational numbers~$x$ and~$y$ such that~$x^y$ is
rational.
\end{prop}
\begin{proof}[First proof] The number~$\sqrt{2}^{\sqrt{2}}$ is rational or
irrational. In the first case, set~$x \defeq \sqrt{2}$, $y \defeq \sqrt{2}$.
In the second case, set~$x \defeq \sqrt{2}^{\sqrt{2}}$, $y \defeq \sqrt{2}$.
\end{proof}
\begin{proof}[Second proof] Set~$x \defeq \sqrt{2}$ and~$y \defeq \log_{\sqrt{2}} 3$.
Then~$x^y = 3$ is rational. The verification that~$y$ is irrational is even
easier than that of~$\sqrt{2}$.\footnote{Let~$y = a/b$ with~$a, b \in \ZZ$
and~$b \neq 0$. Since~$y > 0$, we may assume~$a, b \in \NN$. Then~$3 =
(\sqrt{2})^{a/b}$, hence~$3^{2b} = 2^a$. This is in contradiction to the
uniqueness of the prime factor decomposition, since the factor~$3$ occurs on
the left but not on the right.}
\end{proof}

@@

In constructive mathematics, we do \emph{not} claim that the principle of
excluded middle is false. \marginpar{\dbend}
Indeed, intuitionistic logic is downwardly compatible
with classical logic (every intuitionistic proof is a fortiori also a classical
proof), and some special instances of the principle of excluded middle are
intuitionistically verifiable (an example is given in
Proposition~\ref{prop:discreteness}). Instead, in constructive mathematics we
merely do not use the principle of excluded middle.

Also, deducing from a statement of the form~``$\exists x \in X\_ \varphi(x)$''
that there actually is an element~$x \in X$ such that~$\varphi(x)$, and then
using this particular element in the rest of an argument, is intuitionistically a
valid logical inference, just as it is in classical logic. The axiom of choice
is unrelated to this kind of proof step, even though they are sometimes
mistaken.

\section{Proof by contradiction vs. proof of a negation}

A rumor about constructive mathematics states that constructively, the term
``contradiction'' would be generally forbidden. This is false. In fact, we need
to distinguish between two distinct figures of proof which are often conflated
in classical informal mathematics:

\begin{enumerate}
\item[1.] ``Assume~$\neg\varphi$. Then $\ldots$, a contradiction.
Hence~$\neg\varphi$ was false and thus~$\varphi$ holds.''
\item[2.] ``Assume~$\psi$. Then $\ldots$, a contradiction. Hence~$\neg\psi$.''
\end{enumerate}

Arguments of the first form are proper proofs by contradiction and hence not
generally acceptable in constructive mathematics. What is established is
only~$\neg(\neg\varphi)$, the impossibility of~$\neg\varphi$; constructively,
this is weaker than a positive affirmation of~$\varphi$. Though there are
situations in which~$\neg\neg\varphi$ implies~$\varphi$, generally it does not
and hence intuitionistically arguments of the first form fail to
estabilish~$\varphi$.

In contrast, arguments of the second kind are fine from an intuitionistic point
of view: They are not proper proofs by contradiction, but proofs of negated
statements. That such proofs are valid follows directly from the definition of
negation (which texts on classical logic often also adopt) as a certain
implication:
\[ \neg\psi \defequiv (\psi \Rightarrow \bot), \]
where~``$\bot$'', pronounced ``bottom'', is \emph{absurdity}, a canonical false
statement. (Informally, absurdity is also written as~``$\lightning$'' or~``$1 = 0$''.)
Hence, to establish~$\neg\psi$, we may give a proof of~$\bot$ under
the assumption~$\psi$, just as we may give a proof of~$\beta$ under the
assumption~$\alpha$ if we want to prove~$\alpha \Rightarrow \beta$.

The two proofs of the following fact from number theory demonstrate the
difference:
\begin{prop}\label{prop:sqrt2}The number~$\sqrt{2}$ is not rational.\end{prop}
\begin{proof}[Proof (only valid classically)]
Assume that the claim is false, that is, that the number~$\sqrt{2}$ is
\emph{not not} rational. Then~$\sqrt{2}$ is rational. Hence there are
integers~$a$ and~$b$ with~$\sqrt{2} = a / b$. Thus~$2b^2 = a^2$. This identity
contradicts the uniqueness of the prime factor decomposition, since the
factor~$2$ occurs an odd number of times on the left, but an even number of
times on the right.
\end{proof}
\begin{proof}[Proof (also valid intuitionistically)]
Assume that the number~$\sqrt{2}$ is rational. Then there are integers~$a$
and~$b$, \ldots, a contradiction.
[The theorem on the uniqueness of the prime factor decomposition admits an
intuitionistic proof.]
\end{proof}

Constructively stronger than the statement that~$\sqrt{2}$ is merely
\emph{not rational} is the statement that for every rational number~$x$
the distance~$|\sqrt{2} - x|$ is positive. This stronger claim admits an
intuitionistic proof as well (Exercise~\ref{ex:sqrt2}).

\section{Constructive meaning of mathematical statements}

By our training in classical mathematics, abstaining from the principle of
excluded middle can feel peculiar, perhaps even outrageous: Isn't it obvious
that every mathematical statement is true or false?

This sense of bewilderment is resolved by observing that \emph{even though
constructive mathematicians use the same logical symbols, they have a slightly
different meaning in mind.} When a constructive mathematician argues for some
statement~$\varphi$, they mean that they have an \emph{explicit witness}
for~$\varphi$. This shift in meaning from the classical interpretation is
elaborated by the \emph{Brouwer--Heyting--Kolmogorov interpretation} sketched
below. It is informal and philosophical in nature, but still a useful guide to
the constructive meaning of mathematical statements.

\subsection*{The Brouwer--Heyting--Kolmogorov interpretation}

The notion of witnesses is compositional in nature and begins with witnesses for \emph{atomic
statements}, those not built from substatements using the logical
connectives~$\wedge,\vee,\Rightarrow$ or the quantifiers~$\forall,\exists$. For
instance, in formal number theory, the atomic statements are of the form
``$s = t$'',
where~$s$ and~$t$ are terms for natural numbers; statements of this form are so
simple that we can check them directly so that witnesses don't need to supply
particular additional information. We hence decree that true atomic statements
have trivial witnesses and that false atomic statements have no witnesses at
all.

Table~\ref{table:bhk} explains what witnesses of compound statements are. For
instance, a witness of a statement of the form
\[ \forall n \? \NN\_ \bigl(\varphi(n) \Rightarrow \psi(n)\bigr) \]
is a rule explaining how, for any natural number~$n : \NN$, witnesses
for~$\varphi(n)$ give rise to witnesses for~$\psi(n)$.

\begin{table}
  \centering
  \small
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{@{}rp{5.9cm}p{6.5cm}@{}}
    \toprule
    & {klassische Logik} & {intuitionistische Logik}
    \\\midrule
    Aussage $\varphi$ & Die Aussage $\varphi$ gilt. & Wir haben Beleg für $\varphi$. \\
    $\bot$ & Es stimmt Falschheit. & Wir haben Beleg für Falschheit. \\
    $\varphi \wedge \psi$ & $\varphi$ und $\psi$ stimmen. & Wir haben Beleg für~$\varphi$ und für~$\psi$. \\
    $\varphi \vee \psi$ & $\varphi$ oder $\psi$ stimmt. & Wir haben Beleg für~$\varphi$ oder für~$\psi$. \\
    $\varphi \Rightarrow \psi$ & Sollte~$\varphi$ stimmen, dann auch~$\psi$. &
    Aus Belegen für~$\varphi$ können wir (gleich\-mä\-ßig) Belege für~$\psi$ konstruieren. \\
    $\neg\varphi$ &
      $\varphi$ stimmt nicht. &
      Es kann keinen Beleg für~$\varphi$ geben. \\
    $\forall x\?X\_ \varphi(x)$ & Für alle $x : X$ stimmt jeweils~$\varphi(x).$ &
      Wir können (gleichmäßig) für alle~$x : X$ Belege für~$\varphi(x)$ konstruieren. \\
    $\exists x\?X\_ \varphi(x)$ & \raggedright Es gibt mindestens ein~$x : X$, für das~$\varphi(x)$
    stimmt. & {\raggedright
      Wir haben ein~$x : X$ zusammen mit Beleg für~$\varphi(x).$} \\
    \bottomrule
  \end{tabular}
  \caption{\label{bhk}Informale rekursive Definition des Belegbegriffs.}
\end{table}

\begin{ex}
According to the Brouwer--Heyting--Kolmogorov interpretation, the principle of
excluded middle states that there is a rule for producing, for every~$\varphi$,
either a witness for~$\varphi$ or a witness for~$\neg\varphi$. The BHK
interpretation does not specify which kind of rule ...
Unter dieser Interpretation meint das Prinzip vom ausgeschlossenen Dritten, dass wir für jede
Aussage Beleg für sie oder ihre Negation haben. Das ist aber offensichtlich
nicht der Fall.
\end{ex}

%\begin{bsp}
Die Interpretation von~$\neg\neg\varphi$ ist, dass es keinen Beleg
für~$\neg\varphi$ gibt. Daraus folgt natürlich noch nicht, dass wir tatsächlich
Beleg für~$\varphi$ haben; gewissermaßen ist eine solche Aussage~$\varphi$ nur
"`potenziell wahr"'.
%\end{bsp}

%\begin{bsp}
Wenn wir wissen, dass sich unser Haustürschlüssel irgendwo in der
Wohnung befinden muss (da wir ihn letzte Nacht verwendet haben, um die Tür
aufzusperren), wir ihn momentan aber nicht finden, so können wir konstruktiv
nur folgende doppelt negierte Aussage vertreten:
\[ \neg\neg (\exists x\_ \text{der Schlüssel befindet sich an Position~$x$})
\]
%\end{bsp}

%\begin{bsp}
Wir stehen im Supermarkt und erinnern uns, dass wir unbedingt
gewisse Zutaten einkaufen müssen. Leider fällt uns momentan keine einzige der
Zutaten mehr ein. Dann können wir zwar die Aussage, dass die Menge der zu
besorgenden Zutaten nicht leer ist, vertreten, nicht jedoch die stärkere
Aussage, dass diese Menge ein Element enthält.
%\end{bsp}

%\begin{bsp}[\cite{sigfpe:katemoss,bbc:katemoss}]
Es war ein Video aufgetaucht, dass Kate Moss beim Konsumieren von Drogen zeigte,
und zwar entweder solche von einem Typ~A oder solche von einem Typ~B. Welcher
Typ aber tatsächlich vorlag, konnte nicht entschieden werden. Also gab es für
keine der beiden Straftaten einen Beleg, Kate Moss wurde daher nicht
strafrechtlich verfolgt.
%\end{bsp}

%\begin{bem}Die Formulierung mit dem generischen \emph{Wir} in Tafel~\ref{bhk}
ist etwas irreführend. Wie auch in klassischer Mathematik hängen
intuitionistische Urteile nicht von \emph{uns} oder anderen Mathematikern ab.
Aussagen, die bisher noch nicht konstruktiv bewiesen wurden, können
durchaus einen (noch unbekannten) Beleg besitzen. Eine genauere Diskussion
findet sich etwa in~\cite[Seite~42f.]{kohlenbach:applprooftheory}. Die
Brouwer--Heyting--Kolmogorov-Interpretation kann im Rahmen der
\emph{Realisierbarkeitstheorie} formalisiert werden~\cite{bauer:realizability};
ohne eine solche Formalisierung gibt es gewisse weitere
Probleme~\cite{artemov:bhk,sanz:piecha:critical-bhk,dalen:bhk}.
%\end{bem}

\subsubsection*{\ldots{} über Berechenbarkeit}

Es gibt noch eine zweite Interpretation, die beim Verständnis der konstruktiven
Sichtweise sehr hilfreich ist:
%\listedmotto{konstruktiv-gdw-computer}{Eine Aussage gilt genau dann konstruktiv, wenn es ein
%Computerprogramm gibt, welches sie in endlicher Zeit bezeugt.}
Etwa ist mit dieser Interpretation klar, dass die formale Aussage
\[ \forall n \in \NN\_ \exists p \geq n\_ \text{$p$ ist eine Primzahl}, \]
eine Formulierung der Unendlichkeit der Primzahlen, auch konstruktiv
stimmt: Denn man kann leicht ein Computerprogramm angeben, das eine natürliche
Zahl~$n$ als Eingabe erwartet und dann, etwa über die Sieb-Methode von
Eratosthenes, eine Primzahl~$p \geq n$ produziert (zusammen mit einem Nachweis,
dass~$p$ tatsächlich prim ist).

Das Motto kann man tatsächlich zu einem formalen Theorem
präzisieren, das ist Gegenstand der gefeierten
Curry--Howard-Korrespondenz (Abschnitt~\ref{curry-howard}).

\section{Exercises}
